A novel physical mechanism for the spontaneous formation of porous crystals via viscoelastic phase separation

### 2016-12-12 crystal-gel/nature/crystal_gel3.tex
Merge John's edits: - added temporarily some sections, so it's easier for us to navigate and understand the flow of the manuscript (we can remove the section names before submitting). - added fig:bond_breaking - insist on the role of initial rearrangements of the network to make nucleation possible

\begin{abstract}
{\bf
Viscoelastic phase separation of colloidal suspensions can be interrupted to form gels either by 
%the glass transition, with the formation of colloidal gels~\cite{poon2002,zaccarelli2007,piazza1994phase,verhaegh1997transient,tanaka1999colloid,lu2008gelation}, or by crystallization, which results in the formation of a crystalline network~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways,sabin2012}. 
glass transition~\cite{poon2002,zaccarelli2007,piazza1994phase,verhaegh1997transient,tanaka1999colloid,lu2008gelation} or by crystallization~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways,sabin2012,zhang2012non}. 
%But while the former process has been studied in detail, the latter has never been observed experimentally at a single-particle level and it is thus lacking a microscopic understanding. 
With a new confocal microscopy protocol, we unveil a novel crystallization pathway to sponge-like porous crystal structures in a colloid-polymer mixture undergoing viscoelastic gas-liquid phase separation. We show that this crystallization pathway leads to a network structure different from the original phase separation pattern. Crystals nucleate inside the liquid network, but grow past it by direct condensation of the gas phase on their surface, i.e. de-sublimation, driving evaporation of nearby liquid. %The latter processes via the gas phase has been overlooked so far. %This process represents the colloidal analogue of the Bergeron process~\cite{glickman2000glossary}, which explains the formation of ice crystals in mixed phase clouds and is at the origin of rain. Our finding gives us full experimental access to the kinetic pathway of this poorly understood phenomenon important for climate science. 
%The key is complex dynamic interplay involving the three phases upon simultaneous occurrence of two non-equilibrium phenomena, phase separation and crystallization. 
We argue that similar crystal-gel states can be formed as the result of the gas-liquid phase separation of any crystallizable components, such as monoatomic and molecular systems, if dynamics of the liquid phase is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification. This mechanism will provide a novel single-step pathway to form nano-porous crystals of metals and semiconductors without dealloying \cite{erlebacher2001evolution}, which may be important for catalytic, optical, sensing, and filtration applications \cite{ding2004nanoporous,ding2009nanoporous, wittstock2010nanoporous}.
%, but also sheds new light on crystal network formation in more complex systems such as magma \cite{philpotts1998role}, biominerals \cite{rousseau2005multiscale}, and foods \cite{deman1987fat}.}
}
\end{abstract}

@@

@@

In this Letter, we show that, for low polymer concentration, the system can undergo a novel crystallization pathway that leads to the formation
of crystalline droplets. These droplets originate from crystal seeds that form inside the liquid branches of the spinodal aggregate of network morphology, and then grow by direct condensation from the gas phase. The process is analogous to the Bergeron process of ice formation in mixed phase clouds, where ice droplets grow at the
expense of the supercooled liquid droplets due to their lower saturated vapour pressure. 
This is speculated to be the primary mechanism for the formation of raindrops. 
Differently from the standard crystal-gel scenario, where crystals only nucleate and grow inside the dense branches of the gel, becoming dynamically arrested, we show that the kinetic
pathways involving the gas phase (predominantly the Bergeron process, but also Ostwald ripening) play a fundamental role. 
We reveal that the final crystal-gel network structure has smoother interface than the ordinary gels formed by vitrification because of this novel crystal growth mechanism.  
We also observe that increasing the strength of the interparticle attraction
(by increasing the polymer concentration), instead of forming crystal gels, the system undergoes the classical gelation process, in which
phase separation is interrupted by vitrification~\cite{verhaegh1997transient,tanaka1999colloid,Royall2007,lu2008gelation}.

@@

@@

%%%%
%Previous simulation results~\cite{tanaka2000,tanaka2007spontaneous,furukawa2010key} have suggested that hydrodynamic interactions play a key role in the process of colloidal gelation. Without hydrodynamic interactions, particles have the tendency to aggregate in compact structures and subsequently form thick network structures. This also limits the possibility to form arrested gel states only at relatively high volume fractions. With hydrodynamic interactions, particles first form a transient gel even at very low volume fractions, and only later they increase the number of nearest neighbours to minimize the energy of the structure. Thus, hydrodynamic interactions lead to the formation of gels that are very far from equilibrium and under a strong thermodynamic driving force towards more stable compact structures. The resulting transition from open to more compact networks occurs only through the breaking of the bonds that have accumulated more stress. The process was called stress-driven ageing~\cite{
%tanaka2007spontaneous}, which is characteristic of viscoelastic phase separation \cite{tanaka2000viscoelastic}. Below we provide the first experimental proof of this process.

%In Fig.~\ref{fig:stress}a, we plot the fraction of liquid and gas particles as a function of time during phase separation at the state points of $\phi\approx 0.13$. We observe that at high polymer concentration ($c_p=1.36$, dashed lines) both the liquid and gas branch change monotonically with time as the phase separation proceeds. At low polymer concentration ($c_p=0.82$, continuous lines), on the other hand, we find a non-monotonic evolution of both the liquid and gas branches, which is signalling a structural reorganization of the network. To prove that this is indeed a stress-driven ageing process, in Fig.~\ref{fig:stress}b we plot the average number of bond-breaking events for the low-polymer concentration state point. The figure clearly shows that the network restructuring follows a peak in the number of bond-breaking events. We have thus found direct evidence of the stress-driven ageing process in colloidal gels that is a direct consequence of the formation of a thin network with the help of hydrodynamic interactions in the first stages of the aggregation process.

@@

%%%

While sharing similar structural properties at early times, the samples show significantly different behaviours in the later stages of the phase separation process.
To examine this in more detail, we plot in Fig.~\ref{fig:phase_separation}c the structure factor for the final stages of the gelation process for all state points.
Calculation of the structure factor is done with the Hanning window function, to minimize boundary effects.
The figure shows that at low wavenumber $q$, the structure factor displays fractal scaling compatible with the Guinier law, $S(q)\sim q^{-D}$.
But a difference in the fractal (or effective) dimension $D$ between the $\phi\approx 0.33$ and $c_p=0.38$ mg/g state point and other state points starts to become visible. In fact, while states
with high polymer concentration retain the exponent $D=2.5(3)$, 
which is the random gel universality class exponent, as we also observed in the early stages of the gelation process (see Fig.~\ref{fig:phase_separation}b),
the state point with low polymer concentration ($\phi\approx 0.33$ and $c_p=0.38$ mg/g) displays the biggest deviation from that exponent. In the figure we also
plot the exponent $D=3$, which corresponds to volume growth, which we will show in the following analysis to be the correct exponent for this state. 
The difference is due to the onset of crystallization in the low-polymer concentration sample. This is already evident in the high $q$ behaviour of
the structure factors (Fig.~\ref{fig:phase_separation}c), where the diffraction planes appear as sharper peaks for $\phi\approx 0.33$ and $c_p=0.38$ mg/g.

@@

While nucleation always occurs inside the liquid branches of the phase-separating network, we suggest here the presence of a new growth channel
which was neither considered nor observed before. In the following we will show that the growth mechanism of the crystal is not only due to filling of the spinodal liquid network, as was observed in Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} for Brownian Dynamics simulations, but also involves 
evaporation of the liquid phase and subsequent direct condensation of the gas phase on the crystalline seeds. 
To confirm the presence of the two processes, we plot in Fig.~\ref{fig:crystals}b the gyration radius of individual crystalline nuclei for the state point 
$\phi\approx 0.33$ and $c_p=0.38$ mg/g. The results show that the crystal growth follows two different scaling laws: at small crystalline sizes it scales with 
the fractal (or effective) dimension close to random percolation ($D=2.53$), while at large sizes it scales as $D=3$, as in compact crystal growth. This demonstrates that, while 
small crystalline nuclei are nucleated inside the phase-separated liquid branches, once they reach the transverse size of the liquid branch, the growth 
follows a volume growth with the formation of isotropic crystal droplets. The same scaling law was suggested in the analysis of the
structure factors, Fig.~\ref{fig:phase_separation}c, but here it is shown directly by the analysis of the gyration radius.

@@

The process of formation of ``crystal gels'' may be generic to many other 
systems. The requirements are (i) the presence of gas-liquid phase separation below the melting point of a crystal, (ii) 
weak or little frustration against crystallization (in our case, the use of monodisperse colloids), 
(iii) dynamical slowing down in a supercooled liquid state, which is necessary to induce viscoelastic phase separation leading to the formation 
of a network structure of the minority liquid phase, and (iv) the degree of supercooling is low enough to avoid a high rate of densification of the liquid phase 
leading to vitrification.  
Many monoatomic and single-component molecular systems can satisfy all these conditions in a certain range of the temperature and pressure. 
This can be seen, for example, by looking at the phase diagram of a Lennard-Jones liquid \cite{lodge1997brownian}, which represents many molecular systems 
without specific directional interactions. For monoatomic systems such as noble metals, condition (iii) may not be satisfied easily. 
To access glassy slow dynamics in a liquid phase, we may need a deep quench at a high pressure. However, we note that even without strong dynamic asymmetry 
due to glassiness, bicontinuous phase separation can take place between 35-85 volume \% of the liquid phase in ordinary gas/liquid phase separation \cite{Onuki2002}; 
and, thus, porous crystalline structures can be formed, although a thin network structure may be difficult to be formed.

@@

>>>>

\begin{abstract}
{\bf
Viscoelastic phase separation of colloidal suspensions can be interrupted to form gels either by 
%the glass transition, with the formation of colloidal gels~\cite{poon2002,zaccarelli2007,piazza1994phase,verhaegh1997transient,tanaka1999colloid,lu2008gelation}, or by crystallization, which results in the formation of a crystalline network~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways,sabin2012}. 
glass transition~\cite{poon2002,zaccarelli2007,piazza1994phase,verhaegh1997transient,tanaka1999colloid,lu2008gelation} or by crystallization~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways,sabin2012,zhang2012non}. 
%But while the former process has been studied in detail, the latter has never been observed experimentally at a single-particle level and it is thus lacking a microscopic understanding. 
With a new confocal microscopy protocol, we unveil a novel crystallization pathway to sponge-like porous structures in a colloid-polymer mixture undergoing viscoelastic gas-liquid phase separation.
\textcolor{blue}{
We follow the entire kinetics of phase separation, from the homogeneous phase, percolation, stress driven aging, and until the formation of porous crystals.}
We show that crystallization leads to a network structure different from the original phase separation pattern. Crystals nucleate inside the liquid network, but grow past it by direct condensation of the gas phase on their surface, i.e. de-sublimation, driving evaporation of nearby liquid. %The latter processes via the gas phase has been overlooked so far. %This process represents the colloidal analogue of the Bergeron process~\cite{glickman2000glossary}, which explains the formation of ice crystals in mixed phase clouds and is at the origin of rain. Our finding gives us full experimental access to the kinetic pathway of this poorly understood phenomenon important for climate science. 
%The key is complex dynamic interplay involving the three phases upon simultaneous occurrence of two non-equilibrium phenomena, phase separation and crystallization. 
We argue that similar crystal-gel states can be formed as the result of the gas-liquid phase separation of any crystallizable components, such as monoatomic and molecular systems, if dynamics of the liquid phase is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification. This mechanism will provide a novel single-step pathway to form nano-porous crystals of metals and semiconductors without dealloying \cite{erlebacher2001evolution}, which may be important for catalytic, optical, sensing, and filtration applications \cite{ding2004nanoporous,ding2009nanoporous, wittstock2010nanoporous}.
%, but also sheds new light on crystal network formation in more complex systems such as magma \cite{philpotts1998role}, biominerals \cite{rousseau2005multiscale}, and foods \cite{deman1987fat}.}
}
\end{abstract}

@@

\subsection{Introduction}

@@

\textcolor{blue}{
In this Letter, we show that, for low polymer concentration, the system can undergo a novel crystallization pathway that leads to the formation
of crystalline droplets. These droplets originate from crystal seeds that form inside the liquid branches of the spinodal aggregate of network morphology, and then grow with three different growth channels.
(i) growth within the dense branches of the gel ;
(ii) Bergeron process, analogous to ice formation in mixed phase clouds, where ice droplets grow at the
expense of the supercooled liquid droplets due to their lower saturated vapour pressure ;
(iii) Ostwald ripening.
Differently from the standard crystal-gel scenario, where crystals only nucleate and grow inside the dense branches of the gel, becoming dynamically arrested, we show that the kinetic pathways involving the gas phase (predominantly the Bergeron process, but also Ostwald ripening) play an important role.
We show that a necessary condition for the formation of crystal gels is the stress-driven aging of the network, that is possible only at low polymer concentration, and which releases the stresses
that build-up during the network formation due to hydrodynamic interactions, allowing for an increase in the number of average neighbours. If the average concentration of particles is high enough, this relaxation triggers the nucleation process.
We reveal that the final crystal-gel network structure has smoother interface than the ordinary gels formed by vitrification because of this novel crystal growth mechanism.  
On the other side, we also observe that increasing the strength of the interparticle attraction
(by increasing the polymer concentration), instead of forming crystal gels, the system undergoes the classical gelation process, in which
phase separation is interrupted by vitrification~\cite{verhaegh1997transient,tanaka1999colloid,Royall2007,lu2008gelation}. 
}

\subsection{Samples}

@@

\subsection{Early times: percolation}

@@

\subsection{Intermediate times: stress-driven aging}

\textcolor{blue}
{
At intermediate times, after percolation has started, hydrodynamic interactions play a fundamental role~\cite{tanaka2000,tanaka2007spontaneous,furukawa2010key}.
Without hydrodynamic interactions, as often assumed in simulations, particles have the tendency to aggregate in compact structures and subsequently form thick network structures. This also limits the possibility to form arrested gel states only at relatively high volume fractions. With hydrodynamic interactions, particles first form a transient gel even at very low volume fractions, and only later they increase the number of nearest neighbours to minimize the energy of the structure. Thus, hydrodynamic interactions lead to the formation of gels that are very far from equilibrium and under a strong thermodynamic driving force towards more stable compact structures. The resulting transition from open to more compact networks occurs through the breaking of the bonds that have accumulated more stress. The process was called stress-driven ageing~\cite{tanaka2007spontaneous}, and is characteristic of viscoelastic phase separation \cite{tanaka2000viscoelastic}. Whether the network undergoes such restructuring or not depends on the strength of the bonds, i.e. the polymer concentration.
In Fig.~\ref{fig:bond_breaking} we plot the bond breaking probability, $p_\text{bond}$, and the average coordination, $\bar{n}$, as a function of the time elapsed since salt injection ($t=0$).
At high polymer concentrations (panels \textbf{a}, \textbf{b}), $p_\text{break}$ decreases monotonically, until a stationaty state is reached where $\bar{n}<5$.
At low polymer concentrations (panels \textbf{c}, \textbf{d}), the initial decay of $p_\text{break}$ to a network with $\bar{n}<5$ is followed by a secondary decay to a more compact network with $\bar{n}>6$. The build up of mechanical stress in the first transient network can relax to a more compact network only when when bonds are weak enough, i.e. at low polymer concentrations.
}

\textcolor{blue}{
It is worth noting that the two-step beheviour of $p_\text{break}$ and $\bar{n}$ is not due to crystallization, as this only starts after the network reorganization. For example, the fraction of crystalline particles for all frames in Fig.~\ref{fig:bond_breaking}d is always below $0.3\%$.
The increase in number of bonds after network reorganization is a necessary condition for crystallization, as we will show in the following, while in its absence the network forms low-density arrested states (gels).
 }
%In Fig.~\ref{fig:stress}a, we plot the fraction of liquid and gas particles as a function of time during phase separation at the state points of $\phi\approx 0.13$. We observe that at high polymer concentration ($c_p=1.36$, dashed lines) both the liquid and gas branch change monotonically with time as the phase separation proceeds. At low polymer concentration ($c_p=0.82$, continuous lines), on the other hand, we find a non-monotonic evolution of both the liquid and gas branches, which is signalling a structural reorganization of the network. To prove that this is indeed a stress-driven ageing process, in Fig.~\ref{fig:stress}b we plot the average number of bond-breaking events for the low-polymer concentration state point. The figure clearly shows that the network restructuring follows a peak in the number of bond-breaking events. We have thus found direct evidence of the stress-driven ageing process in colloidal gels that is a direct consequence of the formation of a thin network with the help of hydrodynamic interactions in the first stages of the aggregation process.

@@

%%%
\subsection{Late times: dynamic-arrest vs crystallization}

The samples show significantly different behaviours in the later stages of the phase separation process.
To examine this in more detail, we plot in Fig.~\ref{fig:phase_separation}c the structure factor for the final stages of the gelation process for all state points.
Calculation of the structure factor is done with the Hanning window function, to minimize boundary effects.
The figure shows that at low wavenumber $q$, the structure factor displays fractal scaling compatible with the Guinier law, $S(q)\sim q^{-D}$.
But a difference in the fractal (or effective) dimension $D$ between the $\phi\approx 0.33$ and $c_p=0.38$ mg/g state point and other state points starts to become visible. In fact, while states
with high polymer concentration retain the exponent $D=2.5(3)$, 
which is the random gel universality class exponent, as we also observed in the early stages of the gelation process (see Fig.~\ref{fig:phase_separation}b),
the state point with low polymer concentration ($\phi\approx 0.33$ and $c_p=0.38$ mg/g) displays the biggest deviation from that exponent. In the figure we also
plot the exponent $D=3$, which corresponds to volume growth, which we will show in the following analysis to be the correct exponent for this state. 
The difference is due to the onset of crystallization in the low-polymer concentration sample. This is already evident in the high $q$ behaviour of
the structure factors (Fig.~\ref{fig:phase_separation}c), where the diffraction planes appear as sharper peaks for $\phi\approx 0.33$ and $c_p=0.38$ mg/g.

@@

\subsection{Crystal-growth channels}

While nucleation always occurs inside the liquid branches of the phase-separating network, crystal-growth can proceed through different growth channels.
In the following we will show that the growth mechanism of the crystal is not only due to filling of the spinodal liquid network, as was observed in Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} for Brownian Dynamics simulations, but also involves 
The Bergeron process and Ostwald ripening. 
In Fig.~\ref{fig:crystals}b the gyration radius of individual crystalline nuclei for the state point 
$\phi\approx 0.33$ and $c_p=0.38$ mg/g. The results show that the crystal growth follows two different scaling laws: at small crystalline sizes it scales with 
the fractal (or effective) dimension close to random percolation ($D=2.53$), while at large sizes it scales as $D=3$, as in compact crystal growth. This demonstrates that, while 
small crystalline nuclei are nucleated inside the phase-separated liquid branches, once they reach the transverse size of the liquid branch, the growth 
follows a volume growth with the formation of isotropic crystal droplets. The same scaling law was suggested in the analysis of the
structure factors, Fig.~\ref{fig:phase_separation}c, but here it is shown directly by the analysis of the gyration radius.

@@

The process of formation of ``crystal gels'' may be generic to many other 
systems. The requirements are (i) the presence of gas-liquid phase separation below the melting point of a crystal, (ii) 
weak or little frustration against crystallization (in our case, the use of monodisperse colloids), 
(iii) dynamical slowing down in a supercooled liquid state, which is necessary to induce viscoelastic phase separation leading to the formation 
of a network structure of the minority liquid phase, and (iv) the degree of supercooling is low enough
\textcolor{blue}
{
to allow bond-breaking events that avoid the vitrification of the liquid phase.  
}
Many monoatomic and single-component molecular systems can satisfy all these conditions in a certain range of the temperature and pressure. 
This can be seen, for example, by looking at the phase diagram of a Lennard-Jones liquid \cite{lodge1997brownian}, which represents many molecular systems 
without specific directional interactions. For monoatomic systems such as noble metals, condition (iii) may not be satisfied easily. 
To access glassy slow dynamics in a liquid phase, we may need a deep quench at a high pressure. However, we note that even without strong dynamic asymmetry 
due to glassiness, bicontinuous phase separation can take place between 35-85 volume \% of the liquid phase in ordinary gas/liquid phase separation \cite{Onuki2002}; 
and, thus, porous crystalline structures can be formed, although a thin network structure may be difficult to be formed.

@@

\begin{figure}
 \centering
 \includegraphics[width=16cm]{./bond_breaking_figure.pdf}
 % bond_breaking_figure.pdf: 1553x1066 pixel, 72dpi, 54.79x37.61 cm, bb=0 0 1553 1066
 \caption{{\bf Stress driven aging.} Time evolution of the average coordination of colloidal particles $\bar{n}$ (black continuous line) and the bond-breaking probability $p_\text{break}$ (red dashed line) for state points $(\phi;c_p)$: {\bf a} $(0.14;1.36)$, {\bf b} $(0.33;1.07)$, {\bf c} $(0,14;0.82)$, {\bf d} $(0.33;0.38)$. $p_\text{break}$ is measured as the probability of bonds breaking in a time interval of $\Delta t=10\,s$. For high polymer concentration, panels {\bf a} and {\bf b}, $p_\text{break}$ decreases monotonically until it reaches a stationary state at long times, with $\bar{n}$ saturating at an average of less than $5$ neighbours. For low polymer concentration, panels {\bf c} and {\bf d}, $p_\text{break}$ follows a similar decay at intermediate times, with $\bar{n}$ less than $5$ neighbours, but which is then followed by a second decay to new configuration, where $\bar{n}$ becomes greater than $5$ neighbours.}
 \label{fig:bond_breaking}
\end{figure}
